Demand Forecasting of individual Probability Density Functions with Machine Learning

### 2020-08-08 LaTex/paper.tex
improved model, adapted description

We now refine the method from sec. \ref{sec:cdf_histo} by comparing (in the sense of higher or lower) the quantiles of the observed events, i.e. the sales records, with specified quantile values. For example, if we wanted to check for the quantile value $0.3$, we would look at the observed quantiles for each prediction and calculate the fraction of these being lower than $0.3$, expecting a fraction of $0.3$ for accurate PDF predictions. In order to judge quality of the full PDF predictions, we calculate these fractions for a range of different specified quantile values. When choosing which quantile to compare it is important to cover the tails of the distribution to make sure that even relatively rare events are estimated correctly by the machine learning algorithm. We call this method quantile profile plot.

@@

In summary, this approach has two major benefits compared to the method discussed in sec. \ref{sec:cdf_histo}: First, by explicitly visualizing several different quantiles, the quantile profile plot reveals which part of the predicted PDF, such as the tails of the probability distribution, are particularly problematic. Second, by showing the dependency from the (arbitrary) variable on the x-axis, the quantile profile plot reveals deviations of the predicted PDF from the acuals for different parts, e.g. specific categories, of that variable. Two examples for this can be found in figures \ref{fig:invquant_mean} and \ref{fig:invquant_dayofweek} in the next section.

\subsection{Quantitative Evaluation of PDF Predictions}

The methods discussed  so far allow a detailed qualitative evaluation of PDF predictions. However, in order to also quantify the quality of the PDF predictions, a metric is required that assesses the difference between the distributions. Therefore, we want to compare the CDF histogram of the predicted PDFs with the expected  uniform distribution and define a prediction accuracy interval between 0 and 1 where an accuracy of 1 indicates that both distributions agree perfectly. Several approaches that measure the difference between two probability distributions are suggested in the literature such as  the first Wasserstein distance \cite{olkin1982}, the Kullback-Leibler divergence \cite{kullback1951}, and the Jensen-Shannon divergence  \cite{dagan1997}.

@@

We use data from a Kaggle online competition about estimating unit sales of Walmart retail goods \cite{kaggle_data} to demonstrate forecasting of demand for different product-location-date combinations in form of full individual PDFs by means of Cyclic Boosting and subsequent qualitative and quantitative evaluation of the PDF predictions. The fields identifying a unique sample of the data set are store (\texttt{store\_id}) and item identifiers (\texttt{item\_id}) as well as date, while the field sales represents the target to predict. In the following, we use data from 2013-01-01 to 2016-05-22 for all 10 available stores and 100 different products (\texttt{FOODS\_3\_500}, ..., \texttt{FOODS\_3\_599}) of the department \texttt{FOODS\_3}. Besides the key and target fields (\texttt{item\_id}, \texttt{store\_id}, \texttt{date}, \texttt{sales}), we also use the fields \texttt{event\_name\_1}, \texttt{event\_type\_1}, \texttt{snap\_CA}, \texttt{snap\_TX}, \texttt{snap\_WI}, \texttt{sell\_price}, and \texttt{list\_price} to build features for our machine learning models.

@@

For our Cyclic Boosting mean estimation model, we include the following features: categorical variables for \texttt{store\_id} and \texttt{item\_id}, several variables describing trend and seasonality (days since beginning of 2013 as linear trend as well as day of week, day of year, month, and week of month), time windows around the events given in the data set (7 days before until 3 days after for Christmas and Easter, and 3 days before until 1 day after for all other events like New Year or Thanksgiving), and a flag denoting a promotion. We also include various two-dimensional combinations of the mentioned features, where one of the two dimensions being \texttt{store\_id} or \texttt{item\_id} to learn characteristics of individual locations and products. Furthermore, we include an exponential price-demand elasticity sub-estimator based on the ratio of reduced and normal price and individualized by \texttt{store\_id}. We plan to report on the inner workings of this method in a future work.

\textcolor{red}{... training and prediction periods ... residual correction ...}

\textcolor{red}{... results: mad, mean(y) ... example time series plot ...}

\begin{figure}
\begin{center}
\includegraphics[width=8cm]{../figures/mean_prediction}
\caption{\label{fig:mean_prediction} \textcolor{red}{time series of mean prediction and actuals for one item in all stores}}
\end{center}
\end{figure}

@@

For our Cyclic Boosting variance estimation model, we include the following features: categorical variables for \texttt{store\_id} and \texttt{item\_id}, a flag denoting a promotion, and the insample prediction of the mean model, and the individual correction factor of the residual correction step described in \ref{sec:example_mean}.

\textcolor{red}{... same training and prediction periods ... no residual correction ...}

\textcolor{red}{... pdf/cdf example for visualization...}

\begin{figure}
\begin{center}
\includegraphics[width=4cm]{../figures/pdf}
\includegraphics[width=4cm]{../figures/cdf}
\caption{\label{fig:pdf_example} \textcolor{red}{PDF/CDF example}}
\end{center}
\end{figure}

@@

\textcolor{red}{...}

\begin{figure}
\begin{center}
\includegraphics[width=8cm]{../figures/cdf_truth}
\caption{\label{fig:cdf_demand} \textcolor{red}{CDF histogram}}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=8cm]{../figures/invquant_yhat_mean}
\caption{\label{fig:invquant_mean} \textcolor{red}{invquant profile for mean predictions of all product-location-day combinations on x-axis}}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=8cm]{../figures/invquant_dayofweek}
\caption{\label{fig:invquant_dayofweek} \textcolor{red}{invquant profile for day of week of one product over all locations and days on x-axis}}
\end{center}
\end{figure}

\textcolor{red}{... quantitative CDF accuracy ...}

>>>>

We now refine the method from sec. \ref{sec:cdf_histo} by comparing (in the sense of higher or lower) the quantiles of the observed events, i.e. the sales records, with specified quantile values. In order to do so, we start again from the predicted values for the mean and variance from which a negative binomial distribution is constructed for each prediction, for example for the predicted demand for a single product sold on a single day in a given sales location. Each of these predicted negative binomial PDFs is then transformed to its CDF. Note that for simplicity, we always refer to the negative binomial model in this description, however, the general approach is valid for any PDF.

Then we compare the actual observed sales value (corrected for censored data if necessary) to different quantiles of the corresponding predicted distribution for each data record and average over a larger data sample. For example, if we wanted to check that the median of the distribution, corresponding to the $0.5$ quantile, is predicted correctly by the machine learning model, we would compare the quantile value $0.5$ to the ratio of quantile values of observed sales records being lower/higher than $0.5$. In other words, in case of the median, 50\% of the ex post observed target values should be observed below the median of the corresponding individual predicted PDF and 50\% above.

In order to judge whether the overall shape of the predicted distributions is predicted correctly, we repeat this procedure for a range of quantiles, for example $q = 0.1, 0.2, \ldots, 0.9$. However, we are free to choose which quantiles to look at, and in specific situations it might be advisable to look at the tails of the distribution in more detail, to make sure that even relatively rare events are estimated correctly by the machine learning algorithm, and add more quantiles for comparison in the region between, say $q = 0.95$ and $q = 0.99$. In the following, we call this method {\em quantile profile plot}.

@@

In summary, this approach has two major benefits compared to the method discussed in sec. \ref{sec:cdf_histo}: First, by explicitly visualizing several different quantiles, the quantile profile plot reveals which part of the predicted PDF, such as the tails of the probability distribution, are particularly problematic. Second, by showing the dependency from the (arbitrary) variable on the x-axis, the quantile profile plot reveals deviations of the predicted PDF from the actuals for different parts, e.g. specific categories, of that variable. Two examples for this can be found in figures \ref{fig:invquant_mean} and \ref{fig:invquant_dayofweek} in the next section.

\subsection{Quantitative Evaluation of PDF Predictions}
\label{sec:cdf_acc}

The methods discussed  so far allow a detailed qualitative evaluation of PDF predictions. However, in order to also quantify the quality of the PDF predictions, a metric is required that assesses the difference between the distributions. Therefore, we want to compare the CDF histogram of the predicted PDFs with the expected  uniform distribution and define a metric in the range between 0 and 1, such that the metric takes the value of 1 when both distributions agree perfectly. Several approaches that measure the difference between two probability distributions are suggested in the literature such as  the first Wasserstein distance \cite{olkin1982}, the Kullback-Leibler divergence \cite{kullback1951}, and the Jensen-Shannon divergence  \cite{dagan1997}.

@@

We use data from a Kaggle online competition about estimating unit sales of Walmart retail goods \cite{kaggle_data} to demonstrate forecasting of demand for different product-location-date combinations in form of full individual PDFs by means of Cyclic Boosting and subsequent qualitative and quantitative evaluation of the PDF predictions. The fields identifying a unique sample of the data set are store (\texttt{store\_id}) and item identifiers (\texttt{item\_id}) as well as date. The target $y$ which we need to predict is the number of sales of a given product in a given store on a specific day, denoted by \texttt{sales}. In the following, we use data from 2013-01-01 to 2016-05-22, that describe the sales of 100 different products (\texttt{FOODS\_3\_500}, ..., \texttt{FOODS\_3\_599}) of the department \texttt{FOODS\_3} in 10 stores. All data before 2016 are used as the training data and the data from 2016 are used as an independent test or validation set. Besides the fields used to identify an individual sales record and the corresponding observed sales value, namely \texttt{item\_id}, \texttt{store\_id}, \texttt{date}, \texttt{sales}, we also use the fields \texttt{event\_name\_1}, \texttt{event\_type\_1}, \texttt{snap\_CA}, \texttt{snap\_TX}, \texttt{snap\_WI}, \texttt{sell\_price}, and \texttt{list\_price} to build features for our machine learning models.

@@

The first model, based on the Cyclic Boosting approach described above, is used to predict the mean of an assumed negative binomial distribution, and we use the following variables as features: categorical variables for \texttt{store\_id} and \texttt{item\_id}, several derived variables that are constructed from the time-series of the sales records describing trend and seasonality (days since beginning of 2013 as linear trend as well as day of week, day of year, month, and week of month), time windows around the events given in the data set (7 days before until 3 days after for Christmas and Easter, and 3 days before until 1 day after for all other events like New Year or Thanksgiving), a flag denoting a promotion, and the ratio of reduced (\texttt{sell\_price}) and normal price (\texttt{list\_price}). We also include various two-dimensional combinations of these features. In these cases, one of the two dimensions is either \texttt{store\_id} or \texttt{item\_id}, allowing the machine learning model to learn characteristics of individual locations and products.

Unlike most state-of-the-art time series forecasting methods, we do not include lagged target information, for example via stacking of exponentially weighted moving average features, in our model. The reason for this is that including such information makes the learning of exogenous effects, e.g. promotions or events, much harder, as the model tends to rely mainly on temporal confounding. Omitting these kind of variables improves the capability of the machine learning model to learn causal dependencies, which in turn improves the explainability of the model as well as the quality of the forecasts for mid- to long-term predictions and rare events. In order to capture recent trends that are not reflected in the exogenous features of the model, we apply an individual residual correction on each of the predictions of the machine learning model, which accounts for deviations between the exponentially weighted moving average (with a recursive smoothing factor of $0.15$) of the predictions and targets of each product-location combination over the corresponding past. To reflect a realistic replenishment scenario, we use a prediction horizon of two days for our example here, i.e. a target lag of two days for the training.

\noindent
Just to give a rough verification for the accuracy of the mean predictions, we report the MAD and MSE averaged over the full used data set in the test period to be $1.65$ and $10.15$, respectively, while the average of the target, i.e. the sales, is $3.28$. \fig \ref{fig:mean_prediction} shows the time series of both predictions and sales summed over all 100 products and 10 stores during the test period.

\begin{figure}
\begin{center}
\includegraphics[width=8cm]{../figures/mean_prediction}
\caption{\label{fig:mean_prediction} Time series of mean prediction and sales in test period summed over all products and stores.}
\end{center}
\end{figure}

@@

The second model, also based on Cyclic Boosting, is used to estimate the variance of the negative binomial distribution. The data are split into training and test set as above and in addition the mean predictions for each individual product-location-date combination are fixed in the variance model as stated by \eqn \eqref{eqn:loss_likelihood}. This effectively means that the mean predictions are created in-sample for the training period using the fully trained and validated model for the mean discussed above.

In this model focusing on the variance, we use a different set of features: categorical variables for \texttt{store\_id} and \texttt{item\_id} and a flag denoting a promotion. An example for the resulting PDF and CDF predictions for item \texttt{FOODS\_3\_586} in store \texttt{WI\_1} on 2016-05-22 is shown in \fig \ref{fig:pdf_example}.

\begin{figure}
\begin{center}
\includegraphics[width=4cm]{../figures/pdf}
\includegraphics[width=4cm]{../figures/cdf}
\caption{\label{fig:pdf_example} Predicted PDF (left) and CDF (right) for a specific product-location-date combination.}
\end{center}
\end{figure}

@@

\fig \ref{fig:cdf_demand} shows the histogram of CDF observations according to the method described in sec. \ref{sec:cdf_histo} for all product-location-day combinations in the test period. As benchmark, we compare the outcome of our negative binomial model to a simpler Poisson assumption, which has only a single model parameter, the mean. Using the same mean predictions for both negative binomial and Poisson model, the negative binomial PDF predictions are much closer to the uniform distribution, which we expect for optimal PDF predictions, than the Poisson PDF predictions, showing the effectiveness of our variance estimation. The only significant deviation of the negative binomial histogram from the uniform distribution can be found in the last bins of CDF values close to $1$. As can be seen when excluding samples with mean predictions lower than $1.0$, this deviation stems from slow-sellers, which are prone to overdispersion.

\begin{figure}
\begin{center}
\includegraphics[width=4cm]{../figures/cdf_truth_nbinom}
\includegraphics[width=4cm]{../figures/cdf_truth_poisson}
\includegraphics[width=4cm]{../figures/cdf_truth_nbinom_larger1}
\includegraphics[width=4cm]{../figures/cdf_truth_poisson_larger1}
\caption{\label{fig:cdf_demand} Histograms of ex post observed target CDF values of the corresponding individual PDF predictions (to be compared to a uniform distribution) for our negative binomial model (left) and a simpler Poisson model for comparison (right), using all product-location-day combinations in the test period (upper two plots). In order to show the effect of slow-sellers, the lower two plots exclude all samples with mean predictions lower than $1.0$.}
\end{center}
\end{figure}

\fig \ref{fig:invquant_mean} and \fig \ref{fig:invquant_dayofweek} show quantile profile plots for different variables on the x-axis, namely mean predictions and day of week, respectively. All product-location-day combinations in the test period are used to generate the statistics in the plots. It can be seen that the different weekdays (from Monday as 0 to Sunday as 6) show slightly different patterns and that there are larger deviations for higher mean predictions (with fewer samples though).

\begin{figure}
\begin{center}
\includegraphics[width=8cm]{../figures/invquant_yhat_mean}
\caption{\label{fig:invquant_mean} Quantile profile plot for mean predictions on the x-axis aggregated over all product-location-day combinations in the test period.}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=8cm]{../figures/invquant_dayofweek}
\caption{\label{fig:invquant_dayofweek} Quantile profile plot for the different days of week on the x-axis (from Monday to Sunday) aggregated over all product-location-day combinations in the test period.}
\end{center}
\end{figure}

The quantitative results for the CDF accuracies of our PDF predictions for the different metrics described in sec. \ref{sec:cdf_acc}, calculated over all product-location-day combinations in the test period, can be found in \tab \ref{tab:cdf:acc}. As benchmark, we compare these to the results of a Poisson model assumption needing only the mean as single parameter. Our negative binomial PDF predictions show a significant improvement over the simpler Poisson model, which uses the same mean predictions, for each of the metrics.

\begin{table}[h!]
\begin{center}
\caption{CDF accuracies for negative binomial and Poisson PDF predictions for different metrics calculated over all product-location-day combinations in the test period.}
\label{tab:cdf:acc}
\begin{tabular}{c|c|c}
 & \textbf{NBD} & \textbf{Poisson} \\
\hline
\textbf{EMD} & 0.965 & 0.852 \\
\textbf{KL\_2} & 0.995 & 0.933 \\
\textbf{KL\_e} & 0.997 & 0.954 \\
\textbf{JSD\_2} & 0.999 & 0.982 \\
\textbf{JSD\_e} & 0.999 & 0.987
\end{tabular}
\end{center}
\end{table}
